\begin{frame}{Conclusion}
    \begin{itemize}
        \item In this thesis the transactional approach for concurrent datastructure design was investigated
        \item For Red-Black Trees, the application of commit-time transactions proved unsuccessful
        \item For Skiplists, both encounter-order and commit-time transactions had optimal scaling properties
        \item As opposed to other papers like [5-6], encounter-order transactions outperformed commit-time transactions by a factor of two on all metrics
        \item Limitations of transactional design is its performance
        \item Further research can investigate other common datastructures
    \end{itemize}
\end{frame}

\begin{frame}{References}
\tiny{[1] Maurice  Herlihy  and  J.  Eliot  B.  Moss.  “Transactional  Memory:  Architectural Support for Lock-Free Data Structures”. In:Proceedings ofthe  20th  Annual  International  Symposium  on  Computer  Architecture.ISCA ’93. San Diego, California, USA: Association for Computing Ma-chinery, 1993, pp. 289–300\\}

\vspace{0.3cm}

\tiny{[2] Nir  Shavit  and  Dan  Touitou.  “Software  Transactional  Memory”.  In:Proceedings  of  the  Fourteenth  Annual  ACM  Symposium  on  Principlesof Distributed Computing. PODC ’95. Ottowa, Ontario, Canada: Asso-ciation for Computing Machinery, 1995, pp. 204–213.isbn: 0897917103.doi:10.1145/224964.224987\\}

\vspace{0.3cm}

\tiny{[3] Robert Ennals. “Software transactional memory should not be obstruction free”. In: In Intel Research Cambridge Tech Report. 2006. doi: 10.1.1.702.1468.}

\vspace{0.3cm}

\tiny{[4] Henri Bal et al. “A Medium-Scale Distributed System for Computer Science Research: Infrastructure for the Long Term”. Computer 49.5 (2016), pp. 54–63. doi: 10.1109/MC.2016.127.}

\vspace{0.3cm}

\tiny{[5] David Dice and Nir Shavit. “What Really Makes Transactions Faster?” TRANSACT06 ACM Workshop (Jan. 2006).}

\vspace{0.3cm}

\tiny{[6] Dave Dice, Ori Shalev, and Nir Shavit. “Transactional Locking II”. In: DISC’06. Stockholm, Sweden: Springer-Verlag, 2006, pp. 194–208}
\end{frame}

\begin{frame}{Thank you for your attention!}
Questions?
\end{frame}